\section{Summary and Outlook}
The work done in this thesis was an attempt to establish a robust method for estimating the mutual information between a bottleneck and a parameter distribution during neural posterior estimation with normalizing flows in $21~\si{cm}$ physics. This has a general application in representation learning, where the estimate of mutual information can be used to evaluate the informativeness of a summary network's output distribution. 

The mutual information between the bottleneck and the parameter distribution was estimated for two cases, once for inference of the reionization history, and once for inference of the simulation parameters. For both cases, ensembles of models were trained for various bottleneck dimensions, for which training times of both models were significantly improved.

Next to errors stemming from training instability and a limited sample size, the source of the most ambiguity is that of model selection, since models can always be hand-picked to reproduce the expected results. It was shown that by selecting models based on their best validation performance, subject to the constraint that they do not overfit (demanding a $5\%$ relative distance of training and validation loss), the estimates of mutual information most closely resemble the expected results.

For the case of inference of the neutral hydrogen fraction, mutual information was shown to saturate for larger bottleneck dimensions, though the general trend of the data did not allow an estimate on an optimal bottleneck dimension. This is likely the case because the distributions being analyzed are high-dimensional and being probed with only $500$ samples, and because training was generally less stable than for the inference of the simulation parameters, the relative ensemble errors being much larger.

Mutual information was also shown to saturate for larger bottleneck dimensions during inference of the simulation parameters, and the relative errors on the estimates were improved in comparison with the previous task. The $5\%$ selection rule was shown to fail in one case, suggesting some improvement might be needed.

If applied to other cases, this method for estimating mutual information is best used when comparing models on a large range of bottleneck dimensions. The resolution required to robustly constrain the final estimates between similar bottleneck dimensions needs a larger dataset or larger ensembles, both of which are expensive. Either way, robust estimates can only be obtained if performed on models which generalize well. Additional improvements might be possible if using an $x\%$ decision rule for selecting models, where $x$ is a hyperparameter further tested to obtain more reliable results. Attempting to implement the method for even more complex parameter and bottleneck distributions will make robust estimates much more difficult. 

A useful feature of this method is that it can easily be applied in other settings, when different summary networks are used. If the underlying flow model is kept the same, it can also be used to compare the difference in informativeness across summary networks. There are many other options for summary networks, and one interesting alternative is that of a network which learns in a self-supervised manner, such as SKATR~\cite{Ayo_2025}. The self-supervised model's outputs are independent of the parameters being inferred and as such, multiple normalizing flows for different inference tasks could be trained on the same summary network. Comparing the informativeness of such a network with supervised models would be an interesting goal for the future.