\section{Implementation}



All other cosmological parameters are fixed to the Planck measurements, assuming flat-
ness and a cosmological constant. This means Ωb = 0.04897, $\sigma_8$ = 0.8102, h= 0.6766, and
ns
= 0.9665.

Training is performed on one NVIDIA A30 GPU.
\red{Need specific values of learning rate etc for both models. Should this go in appendix?}

\subsection{The dataset}

The dataset consists of 5000 lightcones simulated with \red{21cmFast, cite the paper somewhere!} where the simulation parameters are drawn from the priors in table~\ref{tab:sim_params}. \red{Add 2018 Planck estimates here instead?}. Each lightcone is originally $140\times 140\times 2350$ voxels in size, corresponding to a redshift range of $z \sim [5,12]$ \red{find out spacial dimensions}. The neutral hydrogen fraction is calculated \red{from the lightcone} and stored as a 92-dimensional vector. The dataset is split into a training, validation and test set with relative sizes $0.7$, $0.2$ and $0.1$, respectively. 

\red{See comments on redshift range in Benedicts paper}

\begin{table}[htbp]
\centering
\caption{Parameters used for the simulations}
\label{tab:sim_params}
\begin{tabular}{p{6cm} l l}

\toprule
Name & Symbol & Prior \\
\midrule
Matter density & $\Omega_m$ & $\mathcal{U}(0.2, 0.4)$ \\
Warm dark matter mass & $\omega_{\text{WDM}}$ & $\mathcal{U}(0.3, 10)~\si{\keV}$ \\
Minimum virial temperature & $T_{\text{vir}}$ & $\mathcal{U}(10^4, 10^{5.3})~\si{\kelvin}$ \\
Ionization efficiency & $\zeta$ & $\mathcal{U}(10, 250)$ \\
Specific X-ray luminosity & $\sigma_8$ & $\mathcal{U}(10^{38}, 10^{42})~\si{\erg\per\second\per\solarMass\times\year}$ \\
X-ray energy threshold for self-absorption by host galaxies & $E_0$ & $\mathcal{U}(100, 1500)~\si{\eV}$  \\

\bottomrule
\end{tabular}
\end{table}

\subsection{The network architecture}

\red{Some introductory sentence...}

Table~\ref{tab:architectures} shows the network architectures of the CNN and the NF for both inference cases \red{Shape values still taken from Caroline's paper, check these still!} The first convolutional layer is used to radically shorten the redshift dimension of the input image, and in turn adding channel dimensions. The next layers gradually reduce the dimensonality in the image's spatial dimensions while adding more channels. Finally, all information is compressed into a 128-dimensonal feature vector, which is transformed into an 6-dimensional output vector. In case the CNN is pretrained, an optional MLP head can be attached.
\begin{table}[htbp]
\centering
\caption{CNN network architecture}
\label{tab:architectures}
\begin{tabular}{l l l}

\toprule
Layer & Image shape \\
\midrule
Input Layer  & (140,140,1916,1) \\
3x3x75 Conv3D  & (138,138,23,32) \\
3x3x2 Conv3D  & (136,136,22,32) \\
2x2x1 Max Pooling  & (68,68,22,32) \\
3x3x2 Conv3D  & (66,66,21,64) \\
1x1x0 Zero Padding  & (68,68,21,64) \\
3x3x2 Conv3D  & (66,66,20,64) \\
2x2x1 Max Pooling  & (33,33,20,64) \\
3x3x2 Conv3D  & (31,31,19,128) \\
1x1x0 Zero Padding  & (33,33,19,128) \\
3x3x2 Conv3D  & (31,31,18,128) \\
Global Average Pooling  & (128) \\
Dense  & (128) \\
Dense  & (128) \\
Dense  & (128) \\
Dense  & (6) \\
\midrule
(Dense)& (128) \\
(Dense)& (128) \\
(Dense)& (N) \\

\bottomrule
\end{tabular}
\end{table}


\subsection{Inference of the neutral hydrogen fraction}

For inference of the neutral hydrogen fraction, the lightcones are preprocessed as in \red{reference yannics paper}: The \red{temperature fluctuation values} are min-max-normalized individually for each lightcone \red{Motivation from CNN?} and the lightcone is truncated to $140\times 140\times 1916$ voxels. This allows the labels to first be truncated to 75 dimensions and then binned to 15-dimensional vectors. This step reduces the dimensionality of the inference problem. \red{Any additional reason for lightcone truncation?} An additional logit transformation is applied to the labels, which improves training performance of the flow. A sigmoid transform can be applied during inference to map the labels back to their physical state. The total dataset has a size of \red{insert current size}.

In order to model the conditional distribution $p(x|y)$, an ensemble of models is trained for each bottleneck dimension. Each ensemble consists of multiple training runs at different weight initializations, i.e. different seeds. The CNN and flow are trained jointly. Both models are trained using an optimizer, a scheduler and a joint scaler. The optimization method is AdamW, the scheduler method is ReduceLROnPlateau and torch's GradScaler ensures more stable backpropagation. A summary of all the hyperparameters and their chosen values can be found in \red{Cite Appendix}. \red{Maybe talk about Optuna?}. 

\red{Talk about significant difficulties with training here, and data methods used to improve training speed. Also mention gradient accumulation.}

Unlike the original distribution of simulation parameters, the distribution of reionization history vectors is not uniform. It therefore has to be modeled separately. It is modeled with a flow of the same architecture as the one from the conditional distribution, without any conditional input. An additional prior model adds uncertainty to the mutual information estimation. The idea behind choosing the same architecture to model the label distribution is that an \red{artefact} of the flow $q_\theta(x|y)$ leading to high variance in the MI estimation might be learned by $q_{\tilde{\theta}}(x)$ as well, thus leading to smaller variance in the final MI estimation.

\subsection{Inference of the simulation parameters}

For inference of the simulation parameters, the data are preprocessed as in \red{reference benedicts paper}: The lightcones are kept in their original dimensionality and lightcones as well as labels are globally standardized.

\red{Talk about adjusting CNN architecture here}

The conditional distribution $p(x|y)$ is again modelled with an ensemble of models for different bottleneck dimensions. Training is done with AdamW optimization and the ReduceLROnPlateau scheduler. No scaler is needed. Using gradient accumulation with an effective batch size of $32$ causes training to fail, therefore gradient accumulation is not implemented. The model does not converge if the CNN and flow are trained jointly from the beginning, so a three-step training process is implemented. First, the CNN is pretrained on the labels with an MSE loss. Since this has to be done for varying bottleneck dimensions, an additional MLP head is added to the existing CNN architecture, mapping the $N_\text{dim}$ output to $6$ dimensions. Pretraining ends after 50 epochs. The CNN weights are then frozen and the MLP head is removed for the rest of training. The CNN outputs after pretraining are then used to pretrain the flow for 120 epochs. Afterwards, the CNN weights are unfrozen and the optimizers and schedulers of both models are reset. Both models are then trained jointly for another 80 epochs. 